---
title: "Chapter 3"
author: "Matt Russell"
date: "8/17/2019"
output: html_document
---

```{r setup, include=FALSE}
#install.packages("nycflights13")
library(nycflights13)
library(tidyverse)
```


```{r}
#Tibbles are data frames for the tidyverse.
nycflights13::flights
View(flights)
```

The five key dplyr functions: filter(), arrange(), select(), mutate(), summarize().

## filter()

Always use == for equals
```{r}
(jan1 <- filter(flights, month==1, day==1))
filter(flights, month == 11 | month == 12)
filter(flights, month %in% c(11,12))
filter(flights, arr_delay <=120, dep_delay <= 120)
filter(flights, !(arr_delay > 120 | dep_delay > 120))
```

## Exercises
```{r}
#1.
filter(flights, arr_delay >= 120)
filter(flights, dest %in% c("IAH","HOU"))
filter(flights, carrier %in% c("UA","AA","DL"))
filter(flights, month %in% c(7,8,9))
filter(flights, arr_delay >= 120 & dep_delay <= 0)
filter(flights, dep_delay >= 60 & arr_delay <= 30)
filter(flights, dep_time >= 0 & dep_time <=600 )

#2. between() is a shortcut for x >= left & x <= right
?dplyr::between
filter(flights, between(dep_time, 0, 600))

#3. 8255 flights have missing dep_time. dep_delay, arr_time, and arr_delay are missing. These are probably canceled flights.
flights.na <- filter(flights, is.na(dep_time))

#4. 
NA^0
```

## arrange()

arrange() changes the orders of rows. NA values are always at the end.
```{r}
arrange(flights, year, month, day)
arrange(flights, desc(arr_delay))
```

## Exercises

```{r}
#1. 
arrange(flights, desc(is.na(arr_delay)))

#2. 
arrange(flights, desc(dep_delay))

arrange(flights, desc(dep_time))

#3.
my.flights <- arrange (flights, air_time)

#4.
my.flights<- arrange(flights, distance) #(Newark to Philly is short)

my.flights<- arrange(flights, desc(distance)) #(JFK to Honolulu is long is long)
```


## select()
select() choose variable from a data frame

```{r}
select(flights, year, month, day)
select(flights, year:day)
select(flights, -(year:day))
rename(flights, tail_num=tailnum)
select(flights, time_hour, air_time, everything())
```

## Exercises
```{r}
View(flights)
#1.
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, dep_time, dep_delay:arr_time, arr_delay)

#2. Adding the same variable multiple times only adds that variable once 
select(flights, dep_time, dep_time)

#3. one_of() selects variables in character vector
?dplyr::one_of()

vars<- c("year", "month", "day", "dep_delay", "arr_delay")

select(flights, one_of(vars))

#4. you can ignore the case. This is supring that the default is case insensitive.
select(flights, contains("TIME"))

select(flights, contains("TIME", ignore.case=FALSE))
```


## mutate()
```{r}
flights_sml<-select(flights,
                    year:day,
                    ends_with("delay"),
                    distance,
                    air_time)
View(flights_sml)

# Mutate makes new variables

mutate(flights_sml,
       gain = arr_delay-dep_delay,
       speed = distance / air_time * 60)

mutate(flights_sml,
        gain = arr_delay-dep_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours)

#transmute() only keeps the new variables you make
transmute(flights,
        gain = arr_delay-dep_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours)  

# Modular arithmetic breaks integers into pieces:

transmute(flights,
          dep_time,
          hour = dep_time %/% 100,
          minute = dep_time %% 100)

# lag() and lead()

(x <- 1:10)
lag(x)
lead(x)

cumsum(x)
cummean(x)

# ranking

y <- c(1,2,2,NA,3,4)

min_rank(y)
min_rank(desc(y))
row_number(y)
```


## Exercises

```{r}
#1. 
flights2<-select(flights,
                    dep_time,
                    sched_dep_time)
mutate(flights2,
       dep_time_min_mdnt = ((dep_time %/% 100)*60)+dep_time %% 100,
       sched_dep_time_min_mdnt = ((sched_dep_time %/% 100)*60)+sched_dep_time %% 100)

#2. you expect to see them the same but they're different
flights2<-select(flights,
                    air_time,
                    arr_time,
                    dep_time)
mutate(flights2,
       air_time2 =arr_time-dep_time,
       arr_time_min_mdnt = ((arr_time %/% 100)*60)+arr_time %% 100,
       dep_time_min_mdnt = ((dep_time %/% 100)*60)+dep_time %% 100,
       air_time3 = arr_time_min_mdnt-dep_time_min_mdnt)

#3. dep_delay should be dep_time-sched_dep_time   
flights3<-select(flights,
                    dep_time,
                    sched_dep_time,
                    dep_delay)
flights3

#4. Greatest delay is 1301 minutes
flights4<-mutate(flights, 
       min_rank(desc(dep_delay)))

arrange(flights4, min_rank(desc(dep_delay)))

View(flights4)

#5. It keeps adding 1, 2, and 3 to 1, 2, 3, 4, ...
1:3 + 1:10

#6. Trigonomic functions include the compute the cosine, sine, tangent, arc-cosine, arc-sine, arc-tangent, and the two-argument arc-tangent.

```

## summarize ()
```{r}

summarize(flights, delay = mean(dep_delay, na.rm=TRUE))
by_day <- group_by(flights, year, month, day)
by_day_summ<-(summarize(by_day, delay = mean(dep_delay, na.rm=TRUE)))

View(by_day_summ)
```

## Piping, example that groups, summarizes, then filters:
```{r}
delays <- flights %>%
  group_by(dest) %>%
  summarize(count = n(),
            dist = mean(distance, na.rm=T),
            delay = mean(arr_delay, na.rm=T)
            ) %>%
  filter(count >20, dest != "HNL")

not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(mean = mean (dep_delay))

delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(delay = mean(arr_delay))

ggplot(data = delays, mapping = aes(x=delay))+
  geom_freqpoly(binwidth = 10)

delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(delay = mean(arr_delay),
            n = n())

#Note the variation decreases as the sample size increases:
ggplot(data=delays, mapping = aes(x=n, y=delay))+
  geom_point(alpha = 0.1)

delays %>%
  filter(n > 25) %>%
  ggplot(mapping = aes(x=n, y=delay))+
  geom_point(alpha = 0.1)
```

## Batting average example

```{r}
#install.packages("Lahman")
library(Lahman)
batting <- as_tibble(Lahman::Batting)

batters <- batting %>%
  group_by(playerID) %>%
  summarize(
    ba = sum(H, na.rm=T)/sum(AB, na.rm=T),
    ab = sum (AB, na.rm=T)
      )

batters %>%
 # filter(ab > 100) %>%
  ggplot(mapping=aes(x=ab, y=ba))+
  geom_point()+geom_smooth(se=F)

batters %>%
  arrange(desc(ba))
```

```{r}
#logical subsetting example:

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(avg_delay1 = mean (arr_delay),
            avg_delay2 = mean (arr_delay[arr_delay > 0]))

not_cancelled %>%
  group_by(dest) %>%
  summarize(sd_distance = sd(distance)) %>%
  arrange(desc(sd_distance))
  
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(first = min (dep_time),
            last = max(dep_time))

# count number of distinct variables 
# *could do this for the number of species on an FIA plot!
not_cancelled %>%
  group_by(dest) %>%
  summarize(carriers = n_distinct(carrier)) %>%
  arrange(desc(carriers))
  
not_cancelled %>%
  count(dest) %>%
  arrange(desc(n))
  
not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(n_early = sum(dep_time < 500))

not_cancelled %>%
  group_by(year, month, day) %>%
  summarize(hour_pct = mean(arr_delay > 60))

daily <- group_by(flights,year, month, day)
(per_day <- summarize(daily, flights=n()))

(per_month<- summarize(per_day, flights=sum(flights)))

(per_year<- summarize(per_month, flights=sum(flights)))

daily %>%
  ungroup() %>%
  summarize(flights=n())

```

# Exercises
```{r}
#1. Arrival delay is more important for the passenger.

#2. 
not_cancelled %>%
  count(dest)

daily <- not_cancelled %>%
group_by(dest)
(per_dest <- summarize(daily, flights=n()))

not_cancelled %>%
  count(tailnum, wt=distance)

daily <- not_cancelled %>%
group_by(tailnum)
(per_tail <- summarize(daily, n=sum(distance)))

#3. Arrival delay because it indicates the amount of delay in arrival.

#4. It seems there are more cancelled flights in the winter
cancelled <- flights %>%
  filter(is.na(dep_delay), is.na(arr_delay))

cancelled %>%
  group_by(year, month, day) %>%
  summarize(n.cancelled = n()) %>%
  arrange(desc(n.cancelled))

#5. Frontier has the longest dep delay (mean value); Hawaian Air had the longest single dep delay (max value)
flights %>%
    group_by(carrier) %>%
  summarize(mean_dep_delay = mean(dep_delay, na.rm=T),
            max_dep_delay = max(dep_delay, na.rm=T),
            n()) %>%
  arrange(desc(max_dep_delay))

# Airlines like Eva Air (EV) seem to have the longest arrival delays
flights %>%
    group_by(carrier, dest) %>%
  summarize(mean_arr_delay = mean(arr_delay, na.rm=T),
            max_arr_delay = max(arr_delay, na.rm=T),
            num_flights = n()) %>%
  filter(num_flights >= 50) %>%
  arrange(desc(mean_arr_delay))

#6. Not sure the question here
flights %>%
    group_by(tailnum) %>%
  summarize(num_flights = n())

# 7.The sort argument to count() sorts the results in order of n. You could use this anytime you would run count() followed by arrange().

```

```{r}
flights_sml %>%
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay))<10)

popular_dests <- flights %>%
  group_by(dest) %>%
  filter(n()>365)

popular_dests %>%
  filter(arr_delay > 0) %>%
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>%
  select(year:day, dest, arr_delay, prop_delay)
```

#Exercises

```{r}
#1. Summary Functions calculate their values within each group when used with mutate() or filter() and group_by().

#2. 
flights %>%
  group_by(carrier, tailnum) %>%
  summarize(mean_arr_delay = mean(arr_delay),
            num_flights = n()) %>%
  filter(num_flights >20) %>%
  arrange(rank(desc(mean_arr_delay)))

my.flights <- flights %>%
    filter(arr_delay > 0) %>%
  group_by(tailnum) %>%
  mutate(prop_delay = arr_delay / sum(arr_delay)) 
  select(tailnum, arr_delay, prop_delay)

flights %>%
  filter(!is.na(tailnum)) %>%
  mutate(on_time = !is.na(arr_time) & (arr_delay <= 0)) %>%
  group_by(tailnum) %>%
  summarise(on_time = mean(on_time), n = n()) %>%
  filter(min_rank(on_time) == 1)

#3. Leave at 7:00 a.m. scheduled dep time!
flights %>%
  mutate(hour_sched = sched_dep_time %/% 100) %>%
  group_by(hour_sched) %>%
  summarize(mean_arr_delay = mean(arr_delay, na.rm=T))

#4. 
flights %>%
  group_by(dest) %>%
  summarize(total_delay = sum(arr_delay, na.rm=T),
            num_flights = n()) %>%
  arrange(rank(desc(total_delay)))

#5. 
lagged_delays <- flights %>%
  arrange(origin, month, day, dep_time) %>%
  group_by(origin) %>%
  mutate(dep_delay_lag = lag(dep_delay)) %>%
  filter(!is.na(dep_delay), !is.na(dep_delay_lag))
lagged_delays

lagged_delays %>%
  group_by(dep_delay_lag) %>%
  summarise(dep_delay_mean = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay_mean, x = dep_delay_lag)) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 1500, by = 120)) +
  labs(y = "Departure Delay", x = "Previous Departure Delay")

#6.
standardized_flights <- flights %>%
  filter(!is.na(air_time)) %>%
  group_by(dest, origin) %>%
  mutate(
    air_time_mean = mean(air_time),
    air_time_sd = sd(air_time),
    n = n()
  ) %>%
  ungroup() %>%
  mutate(air_time_standard = (air_time - air_time_mean) / (air_time_sd + 1))

#7.
flights.dest<- flights %>%
  group_by(dest, carrier) %>%
  summarize (n())

flights.dest %>%
  group_by(dest) %>%
  summarize (num_flights=n()) %>%
  filter(num_flights>=2)
```





